Towards a Pandemic-Scale Ancestral Recombination Graph of SARS-CoV-2

### 2023-05-05 paper.tex
First pass through "Inferred ARGs" section done

\section{Results}
\subsection{Inferred ARGs}
For this preliminary study our goals are to illustrate the utility of
\texttt{sc2ts}, but to also closely study the properties of the
inferred ARGs to inform subsequent development. Thus, rather
than attempting to be comprehensive we chose to work with a
representative subset of the available data, and limited to
inferences that could be performed on a single server in
a few weeks (see below for further details on timings and
computer hardware used). The cut-off dates for sampling were chosen
arbitrarily. We anticipate that further development of the
inference methods and scaling out across multiple servers will
enable inferences at the full pandemic scale using \texttt{sc2ts}
(see Discussion).

To investigate the properties of the method under different
sampling conditions, we inferred two ARGs, which we refer to as the
``Wide'' and ``Long'' ARGs throughout.
The Wide ARG is densely sampled but time-limited and
includes all 1.27 million sample genomes  with a
maximum sampling delay of 30 days (see Methods Section~\ref{sec:filtering_time_travellers})
up to June 30, 2021.
The Long ARG is more sparsely sampled but
includes data for an additional year (June 30, 2022).
For the long ARG, we use at most 1,000 randomly subsampled
genomes per day from those meeting the maximum sampling delay criterion.

@@

    Nodes & & 1,453,347 & & 783,231  & \\
    & Samples & 1,265,685 & (87.09\%) & 657,239 & (83.91\%) \\
    & Local (UPGMA) & 102,709 & (7.07\%) & 51,807 & (6.61\%)  \\
    & Reversion push & 40,538 & (2.79\%) & 34,358 & (4.39\%)  \\
    & Mutation collapse & 40,292 & (2.77\%) & 37,749 & (4.82\%)  \\
    & Recombination & 4,123 & (0.28\%) & 2,078 & (0.27\%) \\
    \cmidrule{2-6}
    % TODO add more information on mutations
    Mutations & & 1.21M &  & 1.06M & \\
    & On one sample & 759K & (63\%) & 767K & (72\%) \\
    & Reversions & 72K & (5.98\%) & 48K & (4.54\%) \\
    & per node & 0.83  &   & 1.36 \\
    \cmidrule{2-6}
    Recombination events & \\
    & Total & 4,123 & & 2,078 & \\
    & Singletons & 2,601  & (63\%) & 1,315 & (63\%) \\
    % clearer for many than "local trees?"
    & Unique breakpoints & 1,496 &  & 958 & \\
    \cmidrule{2-6}
    \multicolumn{2}{l}{Size (inc metadata)} & 57MB & & 36MB&  \\
    \multicolumn{2}{l}{Bytes/sample (exc metadata)}  & 8.29  & & 10.83 \\
    Load time & & 0.9s & & 0.5s & \\
    \bottomrule
\end{tabular}
\end{center}

\caption{\label{tab:args}Summary of the inferred ARGs.
}
\end{table}

% \centering
% \begin{tabular}{l|l|l} \hline \textbf{Property} &
% \textbf{Wide ARG} & \textbf{Long ARG} \\ \hline Last sampling date & 2021-06-30
% & 2022-06-30 \\ Local trees & 1,496 & 958 \\ \hline
% \multicolumn{3}{l}{\textit{Nodes}} \\ All nodes (sample and non-sample) &
% 1,453,347 & 783,231 \\ Sample nodes & 1,265,685 (87\%) & 657,239 (84\%) \\ Tree
% adjustment nodes (excluding recombination nodes) & 80,830 (6.34\%) & 72,107
% (10.97\%) \\ Recombination nodes & 4,123 (0.28\%) & 2,078 (0.27\%) \\ \hline
% \multicolumn{3}{l}{\textit{Types of recombination nodes}} \\ Singletons  & 2,601
% (63\%) & 1,315 (63\%) \\ Two-parent recombinants & 3,521 (85\%) & 1,869 (90\%)
% \\ HMM-consistent recombinants & XX (XX\%) & XX (XX\%) \\ \hline
% \multicolumn{3}{l}{\textit{Mutations}} \\ Total mutations & 1,213,193 &
% 1,062,072 \\ Mutations on a terminal edge & 758,903 (62.55\%) & 767,111
% (72.23\%) \\ Recurrent mutations & 74,719 (6.16\%) & 50,099 (4.72\%) \\
% Reversions & 72,617 (5.99\%) & 48,226 (4.54\%) \\ Immediate reversions (one
% intervening node only) & 3,541 (0.29\%) & 1,347 (0.13\%) \\ \hline
% \multicolumn{3}{l}{\textit{Storage and loading time}} \\ tszipped file
%  & ~57 Mb & ~36 Mb \\ Bytes to encode one
% sample & 8.29 & 10.83 \\ Load time (seconds) & 0.9 & 0.5 \\ \hline
% \end{tabular}

The properties of the inferred ARGs are summarised in Table~\ref{tab:args}.
% TODO more discussion here, making sure we explain anything that might not be
% obvious.
Most of the nodes in the ARGs represent sample genomes
Only a small fraction of the nodes represent recombination events (0.28\% and
0.27\% in the wide ARG and long ARG, respectively), over half of which have no
descendant samples (besides the recombinant sequence that triggered insertion
of the recombination node). There are over one million mutations on
the ARGs, the majority of which occur on a terminal edge. Notably,
the massive amount of genomic variation data in the wide ARG (1.27 million
genomes) and long ARG (657 thousand genomes) are compactly stored in ~57 Mb and
~36 Mb, respectively (less than ten bytes per sample). These ARGs can be loaded
for analysis in a Jupyter notebook on a typical laptop in a
fraction of a second using the Python package tskit.

Inferring the wide ARG required 17 days of computation time on a machine with
128 threads and 512 GB RAM (2x AMD EPYC 7502 @ 2.5GHz). Inferring the long ARG
required 23 days on a (much older) machine with 40 threads and 256 GB RAM (2x
Intel(R) Xeon(R) CPU E5-2680 v2 @ 2.80GHz). The majority of the time is spent
on running the HMM to find the copying path for each sequence, and the process
is therefore highly amenable to distributing across multiple machines.

@@

\subsection{Mutational spectrum}
 In addition to recombination events, the ARGs record mutation events along the
SARS-CoV-2 phylogeny. \cite{Yi2021-sc} reconstructed a SARS-CoV-2 phylogeny of
over 350,000 genomes sampled globally from December 24, 2019, to January 12,
2021, and systematically catalogued all the mutations occurring along the
phylogeny. Notably, they described a mutational spectrum in which C-to-U and
G-to-U mutations occur more frequently than U-to-C and U-to-G mutations,
respectively.

@@

\subsection{Early recombinants reported by Jackson et al. (2021)}
% - Early Recombinants from Jackson et al
%     Discuss results from Wide ARG in detail. Go into the details of the
% forward-backward HMM runs for each example. Be much more explicity about the
% single origin for the different groups (maybe give one of Ana's graphs for
% Group A (XA I think?)

@@

\subsection{Distribution of recombination breakpoint intervals}

@@

\subsection{Divergence between recombinant parents}
% - Divergence between recombinant parents
%     We study the recombination events in the Long ARG in terms. A recombinant
% will not always have two parents, for reasons X, Y and Z. As a first
% approximation, we can assume that each breakpoint corresponds to a
% recombination event (although there are some exceptions). Here we have to get
% into all the hairy details about which sets of breakpoints we filter and don't
% filter.

@@

\subsection{Recombinant Pango lineages}

@@

\subsection{Ancestral Recombination Graphs}
The term ``Ancestral Recombination Graph'' was introduced by
Griffiths and colleagues~\citep{Griffiths1991-two,Griffiths1998-ancestral}
and originally defined an alternative formulation of the coalescent
with recombination stochastic process~\citep{Hudson1983-properties}.
Subsequently, the term ARG came to be used in a more general way to
describe not just realisations of this model, but to any
recombinant genetic ancestry~\citep{Minichiello2006-mapping}.
While there is some subtlety in the details~\citep{Wong2023-efficient},
we can think of an ARG as being any graph that encodes the
reticulate genetic ancestry of a sample of sequences under
the influence of recombination.

@@

\subsection{Node dating}
The approaches to assigning a date to nodes in \texttt{sc2ts} is currently
ad-hoc, and the inferred timing of events from the ARGs reported
here should be treated with caution (e.g., Figure~\ref{fig:recomb_mrcas}).
Sample nodes (those corresponding to observed sample sequences) are the
most accurately dated, as we use the reported collection date for
these nodes. These are not entirely accurate, but our data filtering
criteria should remove the most egregious errors (see
Section~\ref{sec:filtering_time_travellers}). Other nodes in the ARG
are dated by splitting the time between the attached samples
and the chosen parent nodes equally (in the case of
trees inferred from daily sample clusters,
Section~\ref{sec:sample-cluster-tree-inference}) or by adding arbitrary
small values when creating new nodes using parsimony rules
(Section~\ref{sec:parsimony-heuristics}). Substantial improvements could
likely be made using simple node dating methods.

\subsection{Data preprocessing}
The findings of this study are
based on sequences and metadata available on GISAID (\url{https://gisaid.org/})
up to Aug 22, 2022 and accessible at
\url{https://doi.org/10.55876/gis8.230329cd} (Supplementary Table 1).
We removed sequences
if they (1) had ambiguous collection dates;
(2) were collected before January 01, 2020 or after June 30, 2022; or
(3) were isolated from a non-human host.
We aligned sequences to the Wuhan-Hu-1/2019 reference sequence
(GenBank: MN908947.3) using Nextclade v2.3.0~\citep{Aksamentov2021-hj} (dataset tag
2022-07-26T12:00:00Z). We also excluded sequences if they had a
``bad'' quality control status
in any of the four Nextclade columns (``qc.missingData.status'',
``qc.mixedSites.status'', ``qc.frameShifts.status'' and `qc.stopCodons.status'').

>>>>

\section{Results}
\subsection{Inferred ARGs}
For this preliminary study our goals are to illustrate the utility of
\texttt{sc2ts}, and to closely study the properties of the
inferred ARGs to inform subsequent development. Rather
than attempting to be comprehensive we chose to work with a
representative subset of the available data, and limited to
inferences that could be performed on a single server in
a few weeks (see below for further details on timings and
computer hardware used). The cut-off dates for sampling were chosen
arbitrarily.
We inferred two ARGs, which we refer to as the
``Wide'' and ``Long'' ARGs throughout.
The Wide ARG is densely sampled but time-limited and
includes all 1.27 million sample genomes  with a
maximum sampling delay of 30 days
(see Section~\ref{sec:filtering_time_travellers})
up to June 30, 2021.
The Long ARG is more sparsely sampled but
includes data for an additional year (June 30, 2022).
For the long ARG, we use at most 1,000 randomly subsampled
genomes per day from those meeting the maximum sampling delay criterion.

@@

    Nodes & & 1,453,347 & & 783,231  & \\
     & \emph{Node type} & \\
    & Samples & 1,265,685 & (87.09\%) & 657,239 & (83.91\%) \\
    & UPGMA   & 102,709 & (7.07\%) & 51,807 & (6.61\%)  \\
    & Reversion push & 40,538 & (2.79\%) & 34,358 & (4.39\%)  \\
    & Mutation collapse & 40,292 & (2.77\%) & 37,749 & (4.82\%)  \\
    & Recombination & 4,123 & (0.28\%) & 2,078 & (0.27\%) \\
    \cmidrule{2-6}
    Edges & & 1,458,146 &  & 785,539 & \\
     & \emph{Parent node type} & \\
    & Samples & 610,729 & (41.88\%) & 319,626 & (40.69\%) \\
    & UPGMA   & 470,545 & (32.27\%) & 156,881 & (19.97\%)  \\
    & Reversion push & 184,608 & (12.66\%) & 144,991 & (18.46\%)  \\
    & Mutation collapse &186,218  & (12.77\%) & 160,833 & (20.47\%)  \\
    & Recombination & 6,046 & (0.41\%) & 3,208 & (0.41\%) \\
    \cmidrule{2-6}
    Mutations & & 1,231,193 &  & 1,062,072 & \\
    & Per node & 0.83  & $\pm$1.40 & 1.36 & $\pm$1.72 \\
    & Per sample & 0.77  & $\pm$1.39 & 1.38 & $\pm$1.77 \\
    & Per site & 41.23  & $\pm$108.16    & 36.10 & $\pm$80.03\\
    \cmidrule{2-6}
    % size as reported by ls -lh
    \multicolumn{2}{l}{Compressed size (inc metadata)} & 58MB & & 37MB&  \\
    \multicolumn{2}{l}{Bytes/sample (exc metadata)}  & 8.29  & & 10.83 \\
    Load time & & 0.9s & & 0.5s & \\
    \bottomrule
\end{tabular}
\end{center}
\caption{\label{tab:args}Summary of the inferred ARGs. Nodes are classified
as either samples or by the inference process which produced them to aid
understanding of the relative importances of these processes to the overall
topology. Similarly, edges are classified by the type of the parent node,
giving some insights into how often these nodes are subsequently
copied from after insertion (although the relationship is more complicated
for UPGMA inserted nodes). The mean and standard deviation ($\pm$)
are reported for the number of mutations per node, sample and site.
}
\end{table}

The properties of the inferred ARGs are summarised in Table~\ref{tab:args}.
% TODO more discussion here, making sure we explain anything that might not be
% obvious.
Most of the nodes in the ARGs represent sample genomes
(Wide ARG: 87\%, Long ARG: 84\%), but the inferred ancestral sequences
represented by the nodes inferred from daily sample clusters
by UPGMA (Section~\ref{sec:sample-cluster-tree-inference})
and parsimony heuristics (Section~\ref{sec:parsimony-heuristics})
contribute significantly.
Table~\ref{tab:args} shows the counts of edges (see
Section~\ref{sec:args} for a technical definition) in the ARGs
categorised by parent node type, and demonstrates that each
of these contributes significantly to the overall topology.
In particular, we can see that approximately 40\% of the edges
in both ARGs involve inheriting directly from sampled genomes.
Both ARGs contain 29,422 sites (\ref{sec:data_preprocessing}),
and a large number of mutations.
The average number of mutations per-site is high, although much
of this may be explained by outlier sites with artefactually high
mutation counts (see Figure~\ref{fig:breakpoint-distribution}).
Despite this, however, the number of mutations per node
is small. In the Wide ARG, for example, we have a mean of 0.77
mutations per sample, demonstrating that new samples are generally
well represented by the it. See Section~\ref{sec:mutation_spectrum}
for more analyis on the patterns of mutations inferred.
Recombination plays a relatively minor role, with $<0.3\%$
of nodes in the ARGs representing inferred recombination events.
We analyse these events in Sections~\ref{sec:jackson_recombs},
\ref{sec:breakpoint_intervals}, \ref{sec:parent_divergence},
and \ref{sec:pango_x_lineages}.

Table~\ref{tab:args} also summarises some of the computational
properties of the inferred ARGs.
The ARGs are encoded as a ``succinct tree sequence'' using
the \texttt{tskit} library, which provides an extensive
suite of operations for constructing and analysing ARGs
(Section~\ref{sec:args}). For exampl, the Wide ARG
which contains complete genomes (with imputed missing data)
for around 1.2 million samples, along with extensive sample
and debugging metadata, requires only 58MB of space (compressed
using the \texttt{tszip} utility). The majority of this
space is used by metadata, which when discarded results in
an encoding that requires an average of only 8.29 bytes per
SARS-CoV-2 genome stored.
Loading these ARGs takes less than a second, and they can be interactively
analysed using Jupyter notebooks on a standard laptop. The majority
of the analyses in this preprint can be carried out in seconds
with the \texttt{tskit} Python API, using a few gigabytes of RAM.

Inferring these ARGs does require substantial computation.
The wide ARG required 17 days of computation time on a machine with
128 threads and 512 GB RAM (2x AMD EPYC 7502 @ 2.5GHz). The long ARG
required 23 days on a (much older) machine with 40 threads and 256 GB RAM (2x
Intel(R) Xeon(R) CPU E5-2680 v2 @ 2.80GHz). The majority of the time is spent
on running the LS HMM (Section~\ref{sec:ls}) to find the copying
path for each sequence, and the process
is therefore highly amenable to distributing across multiple machines.
We therefore anticipate that further development of the
inference methods and scaling out across multiple servers will
enable inferences at the full pandemic scale
(see Discussion).

@@

\subsection{Mutational spectrum}
\label{sec:mutation_spectrum}
 In addition to recombination events, the ARGs record mutation events along the
SARS-CoV-2 phylogeny. \cite{Yi2021-sc} reconstructed a SARS-CoV-2 phylogeny of
over 350,000 genomes sampled globally from December 24, 2019, to January 12,
2021, and systematically catalogued all the mutations occurring along the
phylogeny. Notably, they described a mutational spectrum in which C-to-U and
G-to-U mutations occur more frequently than U-to-C and U-to-G mutations,
respectively.

@@

\subsection{Early recombinants reported by Jackson et al. (2021)}
\label{sec:jackson_recombs}
% - Early Recombinants from Jackson et al
%     Discuss results from Wide ARG in detail. Go into the details of the
% forward-backward HMM runs for each example. Be much more explicity about the
% single origin for the different groups (maybe give one of Ana's graphs for
% Group A (XA I think?)

@@

\subsection{Distribution of recombination breakpoint intervals}
\label{sec:breakpoint_intervals}

@@

\subsection{Divergence between recombinant parents}
\label{sec:parent_divergence}
% - Divergence between recombinant parents
%     We study the recombination events in the Long ARG in terms. A recombinant
% will not always have two parents, for reasons X, Y and Z. As a first
% approximation, we can assume that each breakpoint corresponds to a
% recombination event (although there are some exceptions). Here we have to get
% into all the hairy details about which sets of breakpoints we filter and don't
% filter.

@@

\subsection{Recombinant Pango lineages}
\label{sec:pango_x_lineages}

@@

\subsection{Ancestral Recombination Graphs}
\label{sec:args}
The term ``Ancestral Recombination Graph'' was introduced by
Griffiths and colleagues~\citep{Griffiths1991-two,Griffiths1998-ancestral}
and originally defined an alternative formulation of the coalescent
with recombination stochastic process~\citep{Hudson1983-properties}.
Subsequently, the term ARG came to be used in a more general way to
describe not just realisations of this model, but to any
recombinant genetic ancestry~\citep{Minichiello2006-mapping}.
While there is some subtlety in the details~\citep{Wong2023-efficient},
we can think of an ARG as being any graph that encodes the
reticulate genetic ancestry of a sample of sequences under
the influence of recombination.

@@

\subsection{Node dating}
\label{sec:node_dating}
The approaches to assigning a date to nodes in \texttt{sc2ts} is currently
ad-hoc, and the inferred timing of events from the ARGs reported
here should be treated with caution (e.g., Figure~\ref{fig:recomb_mrcas}).
Sample nodes (those corresponding to observed sample sequences) are the
most accurately dated, as we use the reported collection date for
these nodes. These are not entirely accurate, but our data filtering
criteria should remove the most egregious errors (see
Section~\ref{sec:filtering_time_travellers}). Other nodes in the ARG
are dated by splitting the time between the attached samples
and the chosen parent nodes equally (in the case of
trees inferred from daily sample clusters,
Section~\ref{sec:sample-cluster-tree-inference}) or by adding arbitrary
small values when creating new nodes using parsimony rules
(Section~\ref{sec:parsimony-heuristics}). Substantial improvements could
likely be made using simple node dating methods.

\subsection{Data preprocessing}
\label{sec:data_preprocessing}
The findings of this study are
based on sequences and metadata available on GISAID (\url{https://gisaid.org/})
up to Aug 22, 2022 and accessible at
\url{https://doi.org/10.55876/gis8.230329cd} (Supplementary Table 1).
We removed sequences
if they (1) had ambiguous collection dates;
(2) were collected before January 01, 2020 or after June 30, 2022; or
(3) were isolated from a non-human host.
We aligned sequences to the Wuhan-Hu-1/2019 reference sequence
(GenBank: MN908947.3) using Nextclade v2.3.0~\citep{Aksamentov2021-hj} (dataset tag
2022-07-26T12:00:00Z). We also excluded sequences if they had a
``bad'' quality control status
in any of the four Nextclade columns (``qc.missingData.status'',
``qc.mixedSites.status'', ``qc.frameShifts.status'' and `qc.stopCodons.status'').
